Hydraulically-vulnerable trees survive on deep-water access in a tropical forest community

### 2020-10-24 manuscript.rmd
1. corrected italics and romal letters in equations and elsewhere 2. added space before paragraphs

---
title: "Hydraulically-vulnerable trees survive on deep-water access in a tropical forest community"
author: ''
date: ''
output:
  bookdown::html_document2:
    number_sections: no
    theme: flatly
    fig_caption: yes
  bookdown::pdf_document2: default
  bookdown::word_document2:
        reference_docx: manuscriptstyle.docx
        fig_caption: yes
bibliography: references.bib
csl: new-phytologist.csl
---

@@

# Introduction
\
Tropical forests play a disproportionately large role in the global carbon and energy cycles [@Bonan:2008], and support half of global biodiversity [@Wright:20050th], but face a threat from intensifying droughts [@Doughty:2015; @Xu.2019]. Tropical forests are logically considered to be an especially drought-vulnerable among biomes given the combination of climate risk and vegetation sensitivity [@Meir:2015trw], yet mortality events in the tropics are rarely as large as those in the temperate and boreal zones (@McDowell.2018), leading to questions regarding the role of large trait-diversity and hydraulic strategies in mitigating mortality events. Despite this, widespread drought-induced tree mortality has occurred across the tropics for specific functional groups [@Bennett:2015; @Chitra-Tarak.2018; @Hilker:2014; @Phillips:2010] and tropical mortality rates are increasing in some regions [@Brienen.2015; @Hubau.2020]. However, state-of-the-art dynamic global vegetation models (DGVMs) struggle to capture drought-induced vegetation dynamics in the tropics [@Galbraith.2010; @Powell:2013; @Powell.2018], as underlying mechanisms of drought-tolerance are not fully understood or quantified.

\     Plants rely on a variety of structural and functional mechanisms to avoid or tolerate a drought, from adjustment of leaf area (deciduousness), change in leaf angle, reductions in stomatal conductance, upregulation of aquaporins, osmotic regulation to stem water storage, deep water access, increased root production, hydraulic redistribution and embolism resistance[@McDowell.2008]. Deep water access is arguably the most long-standing, yet under-studied mechanism. This mechanism assumes that plant available water varies with depth, and that the rooting depths of co-located trees also vary, by species and size [@Meinzer:1999; @Chitra-Tarak.2018; @Brum.2019]. Consequently, trees in the same forest may differ in their experience during a drought, and thus in their growth and mortality responses [@Chitra-Tarak.2018]. A key bottleneck in community-wide testing of this mechanism has been a lack of data in both trees' rooting or water-sourcing depths and plant available soil water at those depths. A recent meta-analysis documented maximum rooting depths for 318 tree species, that is, less than 0.5% of the world’s more than 60 thousand tree species [@Fan:2017; @Beech:2017]. Further, only a small fraction of those are tropical, even though more than 90% of the world’s tree diversity resides in the tropics [@Slik:2015]. Use of stable isotopes of water as a tracer provides an indirect measure of water sourcing depths by matching the isotopic value in xylem water to variations in soil pore water with depth. However, such data are rare [@Evaristo:2016] and their interpretation is an active area of research given unresolved issues pertaining to soil heterogeneity and the correspondence of soil water extract in the laboratory to actual plant available water [@Adams.2020; @Bowers.2020; @Deurwaerder.2020]. DNA barcoding of roots [@Jones.2011] may be used to to estimate species-specific rooting depths or rooting profiles, although extent of sampling required may be cost-prohibitive and DNA barcode libraries for tropical forests are still under development. In general, community-scale data collection for rooting or water-sourcing depths is challenging for species-rich tropical forests.

\     Trade-offs in plant 'traits' shape how plants make a living in their environment leading to alternative ecological strategies. Characterizing the essential constraints necessary to model species-rich communities under a changing environment [@Koven:2020; @Christoffersen:2016kd2d; @Wright:2010; @Bartlett.2019] entails identifying the topography of trait trade-offs in different environments, and how these relate to demographic rates (growth, recruitment and mortality). Significant efforts have been invested into identifying and linking universal drought indices, above-ground traits, and demographic rates. These efforts have resulted in identification of correlations between aboveground hydraulic traits such as vulnerability to embolism and hydraulic safety margins and mortality [@Anderegg:2016], although numerous counter examples also exist [@Hoffmann.2011; @Johnson:2018fad; @Nardini:2015; @Paddock:2013, @Venturas.2016]. Deep roots may mitigate hydraulic vulnerability and mortality-risk, yet plants' realized water environments, specifically their ability to acquire water via differing rooting depths, are rarely studied, especially in the tropical region. We tested how rooting depth is linked to above-ground hydraulic traits and mortality for co-occurring tree species of a tropical forest over 35 years that witnessed droughts of various intensity and duration.

\     We used a locally parameterized and calibrated 1-D hydrological water balance model to obtain daily dynamics of soil water potential by depth in the whole-soil column and then applied a novel inverse model to estimate species effective rooting depths (ERD), defined as the depth at which soil water potential best explained observed dynamics of species-specific growth for large trees of `r length(erd.sp)` dominant species (Fig. \@ref(fig:schematic)). This model incorporated the impact of atmospheric drought on growth and was constrained by above-ground hydraulic traits, namely, species-specific leaf vulnerability curves, by leveraging existing data for eight species and developing trait-based proxies for the rest. We evaluated ERD against existing dueterium to hydrogen content data for a subset of six tree species as an independent observation of ERD, given the corresponding gradient in isotopic composition of water in the soil profile. We tested coordination of ERD with hydraulic traits for a subset of eight species with both datasets. We finally analysed species' mortality rates across the censuses spanning 35 years to test whether ERD explained risk of drought-induced mortality in periods of extreme hydrological drought. Specifically, we tested an overarching hypothesis that inter-specific variation in trees' effective rooting depths (ERD) enables increasingly reliable water availability with depth and thus mitigation of hydraulic vulnerability. Our hypotheses are that (i) deep rooted trees have hydraulic traits associated with rapid water transport but cavitation-vulnerable xylem, whereas shallow rooted species are associated with less efficient but safer hydraulic traits (see Table \@ref(tab:hypo)); and that, (ii) deep rooted species have lower mortality rates due to low realized risk of hydraulic failure, whereas shallow rooted species may have higher mortality rates as a result of higher variability in water availability in shallow soil layers. To our knowledge this is the first study to test for a mechanistic link between plant available water in the whole-soil column, trees' above- and below-ground hydraulic architecture and long-term mortality outcomes for a species-rich forest.

Fig. \@ref(fig:schematic) here.
Table \@ref(tab:hypo) here.

# Materials and Methods
\
In the following sections we describe the study site, the models and data used to estimate and evaluate (1) soil water potential dynamics by depth in the whole-soil column and (2) species-specific effective rooting depths. These are followed by analyses to address the proposed hypotheses. Details of all datasets used are described in the Supplementary Information.

## Study site description
\
We conducted this study at the seasonally dry tropical forest on Barro Colorado Island (BCI), Panama, given its extensive characterization in terms of long-term forest demography, plant functional ecology, hydrology, and data on stable water isotopes on tree species and soil from an independent study of tree water-sourcing depths [@Meinzer:1999]. The 1,560ha island was formed after the river Chagres was dammed in 1914 to create Gatun Lake and complete the Panama Canal. The entire island is forested and classified as a tropical moist forest in the Holdridge Life Zone system. Long-term hydrological monitoring at BCI began in 1972, while demographical monitoring began with the 50 ha ForestGEO plot establishment in 1981-82 [@Condit.1998uq; @Hubbell:1999; @Anderson-Teixeira:2014]. Rainfall at BCI is seasonal with a mean annual rainfall of `r round(rain.man.stats$rain)` mm ($\pm$ `r round(rain.man.stats$rain.sd)` SD; `r min(clim$Year)`-`r max(clim$Year)`) and a four-month long dry season from mid-December through mid-April ($\le$ 100 mm rain month^-1^). Among the tree *species* found in the forest canopy, 28% are identified to be at least partly deciduous during the dry season, but this amounts to a maximum of 6.3% *trees* in the canopy that are deciduous, and the percentage of total forest crown area that is deciduous is estimated at 9.7% [@Condit:2000].

\     In the 50 ha plot all stems above 10 mm diameter at breast height (dbh) are mapped, tagged with a unique number, identified to species and measured every 5 years through 1985-2015 for growth, mortality as well as recruitment of new stems into the 10 mm dbh size class [@Condit.2017]. This inventory represents 321 woody species across the eight censuses. Aside from a 2ha patch of ~100-year-old secondary forest at the north-eastern corner of the plot the remaining 48ha represent old growth forest that was never clear cut for agriculture. The plot (1000 m $\times$ 500 m) was originally chosen for its relative uniformity of relief and other physical conditions; the elevation range is only 40 m (120-160 m) [@Harms:2001]. The plot is situated in the Conrad catchment. Soils of the plot are homogeneous with red light clays, described as the Ava soil type, accounting for 72.4% of the plot [@Baillie.2017]. The topsoil field texture is silty clay loam that gradually fines to silty clay in the subsoil. Soil is mostly free draining, but restricted subsoil permeability gives rise to temporary wet season ponding. Detailed descriptions of the climate, geology, flora and fauna of BCI can be found elsewhere [@Croat.1978; @Leigh.1982; @Gentry.1990].

@@

### Model description
\
To estimate dynamics of water availability by depth over the forest’s rooting zone, we used the Energy Exascale Earth System Land Model (ELM), coupled with the Functionally Assembled Terrestrial Ecosystem Simulator (FATES) (hereafter, ELM-FATES). ELM is a land-model that, among many features, simulates the physics and conservative dynamics of water, energy and carbon fluxes. In particular, soil hydrologic fluxes are vertically resolved among discrete soil layers (1-D) [@Bonan:2013]. FATES is a community-based, open-source model used for studying climate-vegetation interactions. FATES is a vegetation demography model, with a size-structured group of plants (cohorts) and successional trajectory-based patches based on the ecosystem demography [@Moorcroft.2001] approach. FATES couples to ELM by a common interface of water and carbon fluxes. Detailed descriptions for ELM and FATES can be found elsewhere [@Fisher.2010; @Fisher.2015; @Koven:2020].

\     ELM-FATES parameterizes interception, throughfall, canopy drip, infiltration, evaporation, surface runoff, subsurface drainage, redistribution within the soil column, and groundwater discharge and recharge so as to simulate changes in canopy water $\Delta W$~can~, surface water $\Delta W$~sfc~, soil water $\Delta W$~liq,z~ at depth z and water in an unconfined aquifer $\Delta W_{a}$ (omitting processes relevant to snow, wetlands or lakes). Conservation of these terms is expressed as follows:

\begin{equation}
\begin{split}
\Delta W_{can} + \Delta W_{sfc} + \sum_{i=1}^{15} \Delta W_{liq,i} + \Delta W_{a} = \left(q_{rain} - E_{v} - E_{g} - q_{over} - q_{h2osfc} - q_{drai}\right)\Delta t
\end{split}
(\#eq:water-balance)
\end{equation}

where, $q$~rain~ is rainfall, $E_{v}$ is transpiration of the forest, $E_{g}$ is ground evaporation, $q$~over~ is surface runoff, $q$~h2osfc~ is runoff from surface water storage, $q$~drai~ is sub-surface drainage (all in mm H~2~O) and $\Delta t$ is the time-step (typically, 1 hour). The soil profile is discretized into 15 exponentially distributed soil layers with depth z, where z $\in$ Z, and Z = {0.01, 0.03, 0.06, 0.12, 0.21, 0.37, 0.62, 1, 1.7, 2.9, 4.7, 7.8, 13, 21, 35} m.

### Model parameterization
\
To parameterize ELM-FATES, we identified eleven parameters relevant for Eq. \@ref(eq:water-balance), determined their ranges based on literature for the study site, when available, else for the tropics (Table \@ref(tab:param)), and ran 5000 simulations using Latin Hypercube Sampling (LHS) of this global parameter space. In the Conrad catchment, existing data on soil hydraulic conductivity for the depths of 12.5 and 60 cm showed a decline of conductivity by depth (@Kinner:2004; Table \@ref(tab:param)). We generated 5000 LHS samples within the 95% CI for the two depths and assumed a linear decline between each paired sample, while for depths < 12.5 cm and > 60 cm, we assumed the same value as that for depths 12.5 and 60 cm, respectively (Fig. S\@ref(fig:ksat-profiles) [Need to fix SI Figs. numbering to begin at 1]). For greater accuracy, instead of the ELM default soil texture-based pedo-transfer functions, we estimated parameters of soil retention curves using existing data for gravimetric water content vs. $\Psi$~soil~ for depths of 0.15, 0.4 and 1 m [@Kupers:2019]. We converted gravimetric water content to volumetric (VWC, $\theta$) using a bulk density value of of 0.8 g cm^-3^ measured at a point location (see below). A @Campbell.1974 empirical equation was fitted for all depths pooled together, as exploratory analyses indicated differences in water retention among depths were nominal,

\begin{equation}
\Psi_{soil} = \Psi_e\left(\frac{\theta}{\theta_{s}}\right)^{-b}
(\#eq:soil-ret-curve)
\end{equation}
where $\Psi_e$ is the air-entry matric potential, $\theta_{s}$ is the saturated volumetric water content, and $b$, an index for soil pore-size distribution, respectively. $\Psi_e$, $\theta_{s}$ and $b$ that best-fit the data were 200 (mm H~2~O), 0.51 (cm^3^ cm^-3^) and 10 (unitless), respectively.

\   We ran ELM-FATES for the 5000-member parameter ensemble with hourly climate drivers from 1985-2018 [@Faybishenko.2020] initialized with the observed stand structure from the 50 ha plot [@Condit.2019a] in a single site mode. As our key interest here was in deriving soil water availability, we ran ELM-FATES in a lower-complexity configuration: static stand structure, and with a single plant functional type (PFT) of evergreen trees. The latter was chosen as only 9.7% of BCI crown area is dry-season deciduous and addition of a dry-deciduous PFT did not significantly alter results (not shown). Holding the stand structure static rather than allowing it to emerge from physiologically-mediated competition between trees removes a set of internal model feedbacks and thus enables a cleaner experimental design where the only differences between model simulations are the direct physiological effects of perturbed model parameters.

### Model calibration
\
We calibrated ELM-FATES over 2012-2018 against three key fluxes and states in Eq. \@ref(eq:water-balance), namely, (i) evapotranspiration (ET) from the eddy flux tower (2012-2017; see [SI](#SIData.matteo)), comparable to $E_{v} + E_{g}$ in Eq. \@ref(eq:water-balance)), (ii) stream discharge $q_{run} from the Conrad catchment, comprabel to $q_{over} + q_{drai}$; 2012-2018; see [SI](#SIData.steveP)) and (iii) soil volumetric water content (VWC or $\theta_{z}) comparable to W_{liq, z}$) from two sources: a long-term (2012-2018) record of VWC from a point location near the eddy flux tower we obtained with a vertical TDR probe over 0-15 cm depth ($\theta_{0-0.15 m}$) and existing 50 ha plot-wide snap-shot measurements of VWC during the dry season of 2015 and 2016 for the depths of 0.15, 0.4 and 1 m with 1299 samples in total covering all soil types and habitats (see [SI](#SIData.VWC) and @Kupers:2019 for details; hereafter, $\theta_{0.15 m}$, $\theta_{0.4 m}$ and $\theta_{1 m}$ refer to plot-wide averages).

For ELM-FATES calibration we calculated an objective function for each of the 5000-member ensemble by equally weighting standardized RMSE between observations and simulations across all fluxes and states. We describe the objective function $O$ as :

@@

\begin{equation}
\begin{split}
\overline{O} = \frac{1}{6}\sum \left(RMSE^{*}(ET,E_{v} + E_{g}) + RMSE^{*}(q_{run},q_{over} + q_{drai}) + RMSE^{*}(\theta_{0-0.15m}, {W_{liq,\overline{z}}}) \\
& + RMSE^{*}(\theta_{0.15m}, W_{liq,`r soil.depths[4]`m}) + RMSE^{*}(\theta_{0.4~m}, W_{liq,`r soil.depths[6]`~m}) + RMSE^{*}(\theta_{1~m}, W_{liq,`r soil.depths[8]`~m})\right)
(\#eq:obj-func)
\end{split}
\end{equation}

based on monthly sums for both ET and $q$~run~ and daily averages for $\theta$s. $W$~liq,~$_{\overline{z}}$ is the average of daily $W$~liq,z~, where z $\in$ Z, and Z = {`r paste(soil.depths[1:4], collapse=", ")`} m. Superscript * indicates that RMSE is standardised between 0 and 1 across the 5000 simulations.

\   Preliminary analyses showed that minimising the objective function $\overline{O}$, that is, simultaneously fitting the model to all states and fluxes did not capture the soil moisture dynamics at 1 m depth well, our key priority for model calibration. To ensure model fit at 1 m depth, we first chose 300 parameter ensemble members that minimised RMSE between observations and simulations for VWC at the depth of 1 m ($RMSE^{*}(\theta_{1~m}, W_{liq,`r soil.depths[8]`~m}) < 0.2$), and among those 300 ensemble members we chose 100 members with the lowest values for the objective function $\overline{O}$. See Table \@ref(tab:param) for the ranges of best-fit values for different parameters.

### Model evaluation
\
We evaluated ELM-FATES by calculating RMSE between simulated and observed long-term daily VWC for the depths of 0.1, 0.4 and 1 m (2016-2018), baesd on a dataset we had left-out during calibration. We obtained these observations from three horizontal time domain reflectometry probes at the depths of 0.1, 0.4 and 1 m at a point location near the eddy flux tower (see [SI](#SIData.VWC)).

## Hydrological droughts
\   We ran ELM-FATES with the chosen 100 ensemble members from 1985-2018, with the first 5 years used for model spin-up. Extreme soil droughts were identified by depth *z* as days that fell below the 5 percentile of mean $\Psi$~soil,*z*~ across all hydrological realisations values at that depth.

>>>>

---
title: "Hydraulically-vulnerable trees survive on deep-water access in a tropical forest community"
author: ''
date: ''
output:
  bookdown::word_document2:
        reference_docx: manuscriptstyle.docx
        fig_caption: yes
  bookdown::pdf_document2: default
  bookdown::html_document2:
    number_sections: no
    theme: flatly
    fig_caption: yes
bibliography: references.bib
csl: new-phytologist.csl
---

@@

# Introduction

Tropical forests play a disproportionately large role in the global carbon and energy cycles [@Bonan:2008], and support half of global biodiversity [@Wright:20050th], but face a threat from intensifying droughts [@Doughty:2015; @Xu.2019]. Tropical forests are logically considered to be an especially drought-vulnerable among biomes given the combination of climate risk and vegetation sensitivity [@Meir:2015trw], yet mortality events in the tropics are rarely as large as those in the temperate and boreal zones (@McDowell.2018), leading to questions regarding the role of large trait-diversity and hydraulic strategies in mitigating mortality events. Despite this, widespread drought-induced tree mortality has occurred across the tropics for specific functional groups [@Bennett:2015; @Chitra-Tarak.2018; @Hilker:2014; @Phillips:2010] and tropical mortality rates are increasing in some regions [@Brienen.2015; @Hubau.2020]. However, state-of-the-art dynamic global vegetation models (DGVMs) struggle to capture drought-induced vegetation dynamics in the tropics [@Galbraith.2010; @Powell:2013; @Powell.2018], as underlying mechanisms of drought-tolerance are not fully understood or quantified.

\
\     Plants rely on a variety of structural and functional mechanisms to avoid or tolerate a drought, from adjustment of leaf area (deciduousness), change in leaf angle, reductions in stomatal conductance, upregulation of aquaporins, osmotic regulation to stem water storage, deep water access, increased root production, hydraulic redistribution and embolism resistance[@McDowell.2008]. Deep water access is arguably the most long-standing, yet under-studied mechanism. This mechanism assumes that plant available water varies with depth, and that the rooting depths of co-located trees also vary, by species and size [@Meinzer:1999; @Chitra-Tarak.2018; @Brum.2019]. Consequently, trees in the same forest may differ in their experience during a drought, and thus in their growth and mortality responses [@Chitra-Tarak.2018]. A key bottleneck in community-wide testing of this mechanism has been a lack of data in both trees' rooting or water-sourcing depths and plant available soil water at those depths. A recent meta-analysis documented maximum rooting depths for 318 tree species, that is, less than 0.5% of the world’s more than 60 thousand tree species [@Fan:2017; @Beech:2017]. Further, only a small fraction of those are tropical, even though more than 90% of the world’s tree diversity resides in the tropics [@Slik:2015]. Use of stable isotopes of water as a tracer provides an indirect measure of water sourcing depths by matching the isotopic value in xylem water to variations in soil pore water with depth. However, such data are rare [@Evaristo:2016] and their interpretation is an active area of research given unresolved issues pertaining to soil heterogeneity and the correspondence of soil water extract in the laboratory to actual plant available water [@Adams.2020; @Bowers.2020; @Deurwaerder.2020]. DNA barcoding of roots [@Jones.2011] may be used to to estimate species-specific rooting depths or rooting profiles, although extent of sampling required may be cost-prohibitive and DNA barcode libraries for tropical forests are still under development. In general, community-scale data collection for rooting or water-sourcing depths is challenging for species-rich tropical forests.

\
\     Trade-offs in plant 'traits' shape how plants make a living in their environment leading to alternative ecological strategies. Characterizing the essential constraints necessary to model species-rich communities under a changing environment [@Koven:2020; @Christoffersen:2016kd2d; @Wright:2010; @Bartlett.2019] entails identifying the topography of trait trade-offs in different environments, and how these relate to demographic rates (growth, recruitment and mortality). Significant efforts have been invested into identifying and linking universal drought indices, above-ground traits, and demographic rates. These efforts have resulted in identification of correlations between aboveground hydraulic traits such as vulnerability to embolism and hydraulic safety margins and mortality [@Anderegg:2016], although numerous counter examples also exist [@Hoffmann.2011; @Johnson:2018fad; @Nardini:2015; @Paddock:2013, @Venturas.2016]. Deep roots may mitigate hydraulic vulnerability and mortality-risk, yet plants' realized water environments, specifically their ability to acquire water via differing rooting depths, are rarely studied, especially in the tropical region. We tested how rooting depth is linked to above-ground hydraulic traits and mortality for co-occurring tree species of a tropical forest over 35 years that witnessed droughts of various intensity and duration.

\
\     We used a locally parameterized and calibrated 1-D hydrological water balance model to obtain daily dynamics of soil water potential by depth in the whole-soil column and then applied a novel inverse model to estimate species effective rooting depths (ERD), defined as the depth at which soil water potential best explained observed dynamics of species-specific growth for large trees of `r length(erd.sp)` dominant species (Fig. \@ref(fig:schematic)). This model incorporated the impact of atmospheric drought on growth and was constrained by above-ground hydraulic traits, namely, species-specific leaf vulnerability curves, by leveraging existing data for eight species and developing trait-based proxies for the rest. We evaluated ERD against existing dueterium to hydrogen content data for a subset of six tree species as an independent observation of ERD, given the corresponding gradient in isotopic composition of water in the soil profile. We tested coordination of ERD with hydraulic traits for a subset of eight species with both datasets. We finally analysed species' mortality rates across the censuses spanning 35 years to test whether ERD explained risk of drought-induced mortality in periods of extreme hydrological drought. Specifically, we tested an overarching hypothesis that inter-specific variation in trees' effective rooting depths (ERD) enables increasingly reliable water availability with depth and thus mitigation of hydraulic vulnerability. Our hypotheses are that (i) deep rooted trees have hydraulic traits associated with rapid water transport but cavitation-vulnerable xylem, whereas shallow rooted species are associated with less efficient but safer hydraulic traits (see Table \@ref(tab:hypo)); and that, (ii) deep rooted species have lower mortality rates due to low realized risk of hydraulic failure, whereas shallow rooted species may have higher mortality rates as a result of higher variability in water availability in shallow soil layers. To our knowledge this is the first study to test for a mechanistic link between plant available water in the whole-soil column, trees' above- and below-ground hydraulic architecture and long-term mortality outcomes for a species-rich forest.

\
Fig. \@ref(fig:schematic) here.
Table \@ref(tab:hypo) here.

# Materials and Methods

In the following sections we describe the study site, the models and data used to estimate and evaluate (1) soil water potential dynamics by depth in the whole-soil column and (2) species-specific effective rooting depths. These are followed by analyses to address the proposed hypotheses. Details of all datasets used are described in the Supplementary Information.

## Study site description

We conducted this study at the seasonally dry tropical forest on Barro Colorado Island (BCI), Panama, given its extensive characterization in terms of long-term forest demography, plant functional ecology, hydrology, and data on stable water isotopes on tree species and soil from an independent study of tree water-sourcing depths [@Meinzer:1999]. The 1,560ha island was formed after the river Chagres was dammed in 1914 to create Gatun Lake and complete the Panama Canal. The entire island is forested and classified as a tropical moist forest in the Holdridge Life Zone system. Long-term hydrological monitoring at BCI began in 1972, while demographical monitoring began with the 50 ha ForestGEO plot establishment in 1981-82 [@Condit.1998uq; @Hubbell:1999; @Anderson-Teixeira:2014]. Rainfall at BCI is seasonal with a mean annual rainfall of `r round(rain.man.stats$rain)` mm ($\pm$ `r round(rain.man.stats$rain.sd)` SD; `r min(clim$Year)`-`r max(clim$Year)`) and a four-month long dry season from mid-December through mid-April ($\le$ 100 mm rain month^-1^). Among the tree *species* found in the forest canopy, 28% are identified to be at least partly deciduous during the dry season, but this amounts to a maximum of 6.3% *trees* in the canopy that are deciduous, and the percentage of total forest crown area that is deciduous is estimated at 9.7% [@Condit:2000].

\
\     In the 50 ha plot all stems above 10 mm diameter at breast height (dbh) are mapped, tagged with a unique number, identified to species and measured every 5 years through 1985-2015 for growth, mortality as well as recruitment of new stems into the 10 mm dbh size class [@Condit.2017]. This inventory represents 321 woody species across the eight censuses. Aside from a 2ha patch of ~100-year-old secondary forest at the north-eastern corner of the plot the remaining 48ha represent old growth forest that was never clear cut for agriculture. The plot (1000 m $\times$ 500 m) was originally chosen for its relative uniformity of relief and other physical conditions; the elevation range is only 40 m (120-160 m) [@Harms:2001]. The plot is situated in the Conrad catchment. Soils of the plot are homogeneous with red light clays, described as the Ava soil type, accounting for 72.4% of the plot [@Baillie.2017]. The topsoil field texture is silty clay loam that gradually fines to silty clay in the subsoil. Soil is mostly free draining, but restricted subsoil permeability gives rise to temporary wet season ponding. Detailed descriptions of the climate, geology, flora and fauna of BCI can be found elsewhere [@Croat.1978; @Leigh.1982; @Gentry.1990].

@@

### Model description

To estimate dynamics of water availability by depth over the forest’s rooting zone, we used the Energy Exascale Earth System Land Model (ELM), coupled with the Functionally Assembled Terrestrial Ecosystem Simulator (FATES) (hereafter, ELM-FATES). ELM is a land-model that, among many features, simulates the physics and conservative dynamics of water, energy and carbon fluxes. In particular, soil hydrologic fluxes are vertically resolved among discrete soil layers (1-D) [@Bonan:2013]. FATES is a community-based, open-source model used for studying climate-vegetation interactions. FATES is a vegetation demography model, with a size-structured group of plants (cohorts) and successional trajectory-based patches based on the ecosystem demography [@Moorcroft.2001] approach. FATES couples to ELM by a common interface of water and carbon fluxes. Detailed descriptions for ELM and FATES can be found elsewhere [@Fisher.2010; @Fisher.2015; @Koven:2020].

\
\     ELM-FATES parameterizes interception, throughfall, canopy drip, infiltration, evaporation, surface runoff, subsurface drainage, redistribution within the soil column, and groundwater discharge and recharge so as to simulate changes in canopy water $\mathrm{\Delta}W_{\textrm{can}}$, surface water $\mathrm{\Delta}W_{\textrm{sfc}}$, soil water $\mathrm{\Delta}W_{\textrm{liq},z}$ at depth $z$ and water in an unconfined aquifer $\mathrm{\Delta} W_{a}$ (omitting processes relevant to snow, wetlands or lakes). Conservation of these terms is expressed as follows:

\begin{equation}
\begin{split}
\mathrm{\Delta}W_\textrm{can} + \mathrm{\Delta}W_\textrm{sfc} + \sum_{i=1}^{15} \mathrm{\Delta}W_{\textrm{liq},i} + \mathrm{\Delta}W_{a} = \left(q_\textrm{rain} - E_{v} - E_{g} - q_\textrm{over} - q_\textrm{h2osfc} - q_\textrm{drai}\right)\mathrm{\Delta}t
\end{split}
(\#eq:water-balance)
\end{equation}
where, $q_{\textrm{rain}}$ is rainfall, $E_{v}$ is transpiration of the forest, $E_{g}$ is ground evaporation, $q_{\textrm{over}}$ is surface runoff, $q_{\textrm{h2osfc}}$ is runoff from surface water storage, $q$~drai~ is sub-surface drainage (all in mm H~2~O) and $\mathrm{\Delta}t$ is the time-step (typically, 1 hour). The soil profile is discretized into 15 exponentially distributed soil layers with depth $z$, where $z$ $\in$ $Z$, and $Z$ = {0.01, 0.03, 0.06, 0.12, 0.21, 0.37, 0.62, 1, 1.7, 2.9, 4.7, 7.8, 13, 21, 35} m.

To parameterize ELM-FATES, we identified eleven parameters relevant for Eq. \@ref(eq:water-balance), determined their ranges based on literature for the study site, when available, else for the tropics (Table \@ref(tab:param)), and ran 5000 simulations using Latin Hypercube Sampling (LHS) of this global parameter space. In the Conrad catchment, existing data on soil hydraulic conductivity for the depths of 12.5 and 60 cm showed a decline of conductivity by depth (@Kinner:2004; Table \@ref(tab:param)). We generated 5000 LHS samples within the 95% CI for the two depths and assumed a linear decline between each paired sample, while for depths < 12.5 cm and > 60 cm, we assumed the same value as that for depths 12.5 and 60 cm, respectively (Fig. S\@ref(fig:ksat-profiles) [Need to fix SI Figs. numbering to begin at 1]). For greater accuracy, instead of the ELM default soil texture-based pedo-transfer functions, we estimated parameters of soil retention curves using existing data for gravimetric water content vs. $\Psi$~soil~ for depths of 0.15, 0.4 and 1 m [@Kupers:2019]. We converted gravimetric water content to volumetric (VWC, $\theta$) using a bulk density value of of 0.8 g cm^-3^ measured at a point location (see below). A @Campbell.1974 empirical equation was fitted for all depths pooled together, as exploratory analyses indicated differences in water retention among depths were nominal,

\begin{equation}
\Psi_\textrm{soil} = \Psi_e\left(\frac{\theta}{\theta_{s}}\right)^{-b}
(\#eq:soil-ret-curve)
\end{equation}
where $\Psi_e$ is the air-entry matric potential, $\theta_{s}$ is the saturated volumetric water content, and $b$, an index for soil pore-size distribution, respectively. $\Psi_e$, $\theta_{s}$ and $b$ that best-fit the data were 200 (mm H~2~O), 0.51 (cm^3^ cm^-3^) and 10 (unitless), respectively.

\
\   We ran ELM-FATES for the 5000-member parameter ensemble with hourly climate drivers from 1985-2018 (Faybishenko et al. 2020) initialized with the observed stand structure from the 50 ha plot [@Condit.2019a] in a single site mode. As our key interest here was in deriving soil water availability, we ran ELM-FATES in a lower-complexity configuration: static stand structure, and with a single plant functional type (PFT) of evergreen trees. The latter was chosen as only 9.7% of BCI crown area is dry-season deciduous and addition of a dry-deciduous PFT did not significantly alter results (not shown). Holding the stand structure static rather than allowing it to emerge from physiologically-mediated competition between trees removes a set of internal model feedbacks and thus enables a cleaner experimental design where the only differences between model simulations are the direct physiological effects of perturbed model parameters.

### Model calibration

We calibrated ELM-FATES over 2012-2018 against three key fluxes and states in Eq. \@ref(eq:water-balance), namely, (i) evapotranspiration (ET) from the eddy flux tower (2012-2017; see [SI](#SIData.matteo)), comparable to $E_{v} + E_{g}$ in Eq. \@ref(eq:water-balance)), (ii) stream discharge $q_\textrm{run}$ from the Conrad catchment, comparabel to $q_\textrm{over} + q_\textrm{drai}$; 2012-2018; see [SI](#SIData.steveP)) and (iii) soil volumetric water content (VWC or $\theta_{z}$) comparable to $W_{\textrm{liq,} z}$ from two sources: a long-term (2012-2018) record of VWC from a point location near the eddy flux tower we obtained with a vertical TDR probe over 0-15 cm depth ($\theta_\textrm{0-0.15 m}$) and existing 50 ha plot-wide snap-shot measurements of VWC during the dry season of 2015 and 2016 for the depths of 0.15, 0.4 and 1 m with 1299 samples in total covering all soil types and habitats (see [SI](#SIData.VWC) and @Kupers:2019 for details; hereafter,  $\theta_{\textrm{0-0.15 m}}$, $\theta_{\textrm{0.4 m}}$ and $\theta_{\textrm{1 m}}$ refer to plot-wide averages).

\
\    For ELM-FATES calibration we calculated an objective function for each of the 5000-member ensemble by equally weighting standardized RMSE between observations and simulations across all fluxes and states. We describe the objective function $O$ as :

@@

\begin{equation}
\begin{split}
\overline{O} = \frac{1}{6}\sum \textrm{RMSE}^{*}\textrm{(ET,}E_{v} + E_{g}) + \textrm{RMSE}^{*}(q_{\textrm{run}},q_{\textrm{over}} + q_{\textrm{drai}}) + \textrm{RMSE}^{*}(\theta_{\textrm{0-0.15~m}}, {W_{\textrm{liq,}\overline{z}}}) \\
& + \textrm{RMSE}^{*}(\theta_{\textrm{0.15~m}}, W_{\textrm{liq,`r soil.depths[4]`m}}) + \textrm{RMSE}^{*}(\theta_{\textrm{0.4~m}}, W_{\textrm{liq,`r soil.depths[6]`~m}}) + \textrm{RMSE}^{*}(\theta_{1~m}, W_{\textrm{liq,`r soil.depths[8]`~m}})
(\#eq:obj-func)
\end{split}
\end{equation}

based on monthly sums for both ET and $q$~run~ and daily averages for $\theta$s. $W_{\textrm{liq,}\overline{z}}$ is the average of daily $W_{\textrm{liq,} z}$, where $z$ $\in$ $Z$, and $Z$ = {`r paste(soil.depths[1:4], collapse=", ")`} m. Superscript * indicates that RMSE is standardised between 0 and 1 across the 5000 simulations.

\   Preliminary analyses showed that minimising the objective function $\overline{O}$, that is, simultaneously fitting the model to all states and fluxes did not capture the soil moisture dynamics at 1 m depth well, our key priority for model calibration. To ensure model fit at 1 m depth, we first chose 300 parameter ensemble members that minimised RMSE between observations and simulations for VWC at the depth of 1 m ($\textrm{RMSE}^{*}(\theta_{\textrm{1 m}}, W_{\textrm{liq,`r soil.depths[8]` m}})$ < 0.2), and among those 300 ensemble members we chose 100 members with the lowest values for the objective function $\overline{O}$. See Table \@ref(tab:param) for the ranges of best-fit values for different parameters.

### Model evaluation

We evaluated ELM-FATES by calculating RMSE between simulated and observed long-term daily VWC for the depths of 0.1, 0.4 and 1 m (2016-2018), baesd on a dataset we had left-out during calibration. We obtained these observations from three horizontal time domain reflectometry probes at the depths of 0.1, 0.4 and 1 m at a point location near the eddy flux tower (see [SI](#SIData.VWC)).

## Hydrological droughts

We ran ELM-FATES with the chosen 100 ensemble members from 1985-2018, with the first 5 years used for model spin-up. Extreme soil droughts were identified by depth $z$ as days that fell below the 5th percentile of $\Psi_{\textrm{soil},z}$ for a given day of the year.

## Model for effective rooting depth (ERD)
